%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  CHAPTER 1: FOUNDATIONS
%  Template for first chapter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Some Mathematical Preliminaries}
\label{ch:foundations}

\section{Introduction}

This chapter introduces some fundamental mathematics that will be needed in the later chapters. The terms appearing in this chapter should be familiar to any student of physics. However, due to the intuitive way these terms are usually presented in any first course of calculus at the university freshman level, students are not always familiar with the more "mathematical" way of presenting these concepts. Since in the later study of basic differential geometry these will work as a necessary prerequisite vocabulary, it is better to be familiar with the language a bit. 


\begin{comment}

\subsection{Overview}

[Write brief overview of chapter contents here]

\section{Core Concepts}

\subsection{Definition 1.1: Your First Definition}

\begin{definition}[Your Definition Name]
    Provide a precise mathematical or conceptual definition here. Use proper notation established in the preamble.
    
    For example: A function $f: \R \to \R$ is continuous at $x_0$ if for every $\eps > 0$, there exists $\delta > 0$ such that $|x - x_0| < \delta$ implies $|f(x) - f(x_0)| < \eps$.
\end{definition}

\begin{example}
    Provide a concrete example that illustrates the definition.
\end{example}

\subsection{Theorem 1.1: Your First Result}

\begin{theorem}[Name of Theorem]
    State the theorem clearly here. 
    
    Suppose conditions (i), (ii), and (iii) hold. Then conclusion follows: $A = B$.
\end{theorem}

\begin{proof}
    Outline the proof strategy, then provide key steps.
    
    Step 1: First, we observe that...
    
    Step 2: By applying [technique], we get...
    
    \begin{equation}
        \mathcal{E} = \int_{\Omega} \rho e \, \dd{V}
        \label{eq:energy_def}
    \end{equation}
    
    Step 3: Combining these results yields the conclusion. \qed
\end{proof}

\begin{remark}
    Important observations or extensions of the theorem can go here. This theorem is particularly useful when...
\end{remark}

\section{Key Mathematical Tools}

\subsection{Notation and Operations}

Let's establish the notation we'll use throughout. We denote:
\begin{itemize}
    \item Vectors: $\bv{u} = (u_1, u_2, u_3)$
    \item Matrices: $\mathbf{A} = \{a_{ij}\}$
    \item Derivatives: $\ddt{u} = \frac{\mathrm{d}u}{\mathrm{d}t}$, $\pd{u}{x} = u_{,x}$
\end{itemize}

\subsection{Important Equations}

An important fundamental equation in this field is:

\begin{equation}
    \boxeq{
        \pd{u}{t} + \nabla \cdot \bv{F}(\bv{u}) = 0
    }
    \label{eq:conservation_law}
\end{equation}

This is a conservation law where:
\begin{itemize}
    \item $u$ is the conserved quantity (scalar or vector)
    \item $\bv{F}(\bv{u})$ is the flux function
    \item The equation states that changes in $u$ equal divergence of the flux
\end{itemize}

Equation \eqref{conservation_law} is fundamental and appears repeatedly throughout the course.

\section{Practical Applications}

\subsection{Application 1: [Your Application]}

Consider a specific example where the concepts apply:

\begin{equation}
    \text{Application 1} \quad \rightarrow \quad \text{Result}
    \label{eq:application1}
\end{equation}

The physical significance is that [explain the meaning].

\subsection{Application 2: [Your Application]}

Another important application involves:

\begin{equation}
    \text{Condition} \quad \Rightarrow \quad \text{Consequence}
    \label{eq:application2}
\end{equation}

This demonstrates how [explain the broader context].

\section{Worked Examples}

\begin{example}[Simple Calculation]
    \label{ex:simple_calc}
    
    Given: $\bv{u} = (1, 2, 3)$ and $\bv{v} = (4, 5, 6)$
    
    Find: $\bv{u} \cdot \bv{v}$
    
    \textit{Solution:}
    \begin{align}
        \bv{u} \cdot \bv{v} &= (1)(4) + (2)(5) + (3)(6) \\
        &= 4 + 10 + 18 \\
        &= 32
    \end{align}
\end{example}

\begin{example}[Complex Problem]
    \label{ex:complex_calc}
    
    This more involved example demonstrates integration of multiple concepts. [Provide detailed walkthrough]
\end{example}

\section{Summary and Key Takeaways}

\begin{itemize}
    \item \textbf{Key Point 1:} [Main concept from section 1]
    \item \textbf{Key Point 2:} [Main concept from section 2]
    \item \textbf{Key Point 3:} [Connection to later material]
\end{itemize}

These foundational concepts will be essential for understanding the more advanced material in subsequent chapters.

\section{Exercises}

\begin{enumerate}
    \item \textbf{Basic:} [Simple computational problem]
    \item \textbf{Intermediate:} [Problem requiring multiple concepts]
    \item \textbf{Advanced:} [Challenge problem extending the theory]
    \item \textbf{Proof:} [Problem asking to prove a related result]
\end{enumerate}

\section{Further Reading}

For deeper understanding of the topics covered in this chapter, consult:
\begin{itemize}
    \item [Reference 1] - Good introduction to [topic]
    \item [Reference 2] - Comprehensive treatment of [topic]
    \item [Reference 3] - Advanced material on [topic]
\end{itemize}

%\newpage  % Uncomment to force new page for next chapter

\end{comment}